Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2020-01-26 rnni_convexity.tex
minor clarificants & wording tweaks

Because the algorithm returns a path between pairs of trees in $\rnni$, the length of the path approximates the $\rnni$ distance from above.
A natural question then is how accurate this approximation is.
We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to seven taxa \autocite{Collienne2019}.

@@

The example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Property does not hold.

@@

\begin{conjecture}
\label{conjecture:weak_split_theorem}
The Weak Split Property holds for $\rnni$.
\end{conjecture}

@@

These considerations motivate us to conjecture the following strong version of the Cluster Property.

@@

\begin{conjecture}
\label{conjecture:cluster_theorem}
The Cluster Property holds for $\rnni$.
\end{conjecture}

We have computationally checked that this conjecture is true in $\rnni$ spaces with up to seven taxa ($n = 2, \ldots, 7$).
Specifically, we computed subgraphs of $\rnni$ containing only trees with shared clusters and compared distances in this subgraph with distances in the whole $\rnni$ graph.
We computed these graphs according to the algorithm from \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
The source code of all these implementations can be openly accessed at \autocite{Collienne2019}.

@@

Our results suggest a surprising advance in this field of $\np$-problems -- adding ranking information to trees simplifies some of the algorithms and makes some unintuitive counterexamples impossible.
We hence continued the investigation of the $\rnni$ graph on ranked phylogenetic trees in this paper.
We have taken the route of comparing the classic $\nni$ with the $\rnni$ graph, so have mostly been concentrating on questions that are settled in $\nni$.
Specifically, we have considered a number of geometric properties, including the diameter, radius, convexity of caterpillar trees, and the Split Property.
We have also justified that the technique of \textcite{Dasgupta2000-xa} to prove that distances in $\nni$ are $\np$-hard to compute cannot be directly applied in $\rnni$.
This led us to the Cluster Property, a statement that does not hold in $\nni$.
All our algorithms and methods developed in this paper witness in favour of the Cluster Property being true in $\rnni$.
For example, we have checked the statement computationally for up to seven taxa \autocite{Collienne2019}.
Although the Split Property is true in $\spr$, it is still $\np$-hard to compute distances in that graph.
Hence Cluster Property alone cannot be used as an argument in favour of the distance problem having an efficient solution in $\rnni$.
But the greedy nature of all our algorithms for computing shortest paths in $\rnni$ developed in this paper does provide such an argument.

>>>>

Because the algorithm returns a path between pairs of trees in $\rnni$, the length of the path approximates the $\rnni$ distance from above.
A natural question then is how accurate this approximation is.
We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to seven taxa \autocite{Collienne2019} and know of no larger counterexample.

@@

The example in Figure~\ref{fig:splitthm_counterexample} shows that, in this form, the Split Property is not present.

@@

\begin{conjecture}
\label{conjecture:weak_split_theorem}
$\rnni$ has the Weak Split Property.
\end{conjecture}

@@

These considerations motivate us to conjecture the following Cluster Property.

@@

\begin{conjecture}
\label{conjecture:cluster_theorem}
$\rnni$ has the Cluster Property.
\end{conjecture}

We have found no counterexamples to this conjecture and have checked that it holds in $\rnni$ spaces with up to seven taxa ($n = 2, \ldots, 7$).
Specifically, we computed subgraphs of $\rnni$ containing only trees with shared clusters and compared distances in this subgraph with distances in the whole $\rnni$ graph.
We computed these graphs according to the algorithm from \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
The source code of all these implementations can be openly accessed at \autocite{Collienne2019}.

@@

Our results suggest a surprising advance in this field of $\np$-problems -- adding ranking information to trees simplifies some of the algorithms and makes some unintuitive counterexamples impossible.
We hence continued the investigation of the $\rnni$ graph on ranked phylogenetic trees in this paper.
We have taken the route of comparing the classic $\nni$ with the $\rnni$ graph, so have mostly been concentrating on questions that are settled in $\nni$.
Specifically, we have considered a number of geometric properties, including the diameter, radius, convexity of caterpillar trees, and the Split Property.
We have also justified that the technique of \textcite{Dasgupta2000-xa} to prove that distances in $\nni$ are $\np$-hard to compute cannot be directly applied in $\rnni$.
This led us to the Cluster Property, a statement that does not hold in $\nni$.
All our algorithms and methods developed in this paper witness in favour of $\rnni$ having the Cluster Property.
For example, we have checked the statement computationally for up to seven taxa \autocite{Collienne2019}.
Although the Split Property is present in $\spr$, it is still $\np$-hard to compute distances in that graph.
Hence the Cluster Property alone cannot be used as an argument in favour of the distance problem having an efficient solution in $\rnni$.
But the greedy nature of all our algorithms for computing shortest paths in $\rnni$ developed in this paper does provide such an argument.
